---
title: "What exactly do numbers mean?"
subtitle: "Analysing our covered-box replication"
author: "LX 433/533/733 — Experimental Pragmatics"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: readable
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2, dpi = 150)
```

```{r libraries}
library(tidyverse)
```

## Before anything else

Two files decide what everything below means, and both are written by the same
scripts that built the survey, so they cannot drift out of step with it:

- **`columns.txt`** — the response columns, in survey order.
- **`choice-map.csv`** — what each choice ID means on each question. Choice IDs
  are stable no matter what order the boxes appeared in, but *what* choice 1 and
  2 stand for varies by trial type, so nothing below hard-codes it.

Set `data_file` to the fake data to rehearse, or to the real export on the day.

```{r data-file}
data_file <- "coveredbox-fake-data.csv"
# data_file <- "CoveredBox_September+3,+2026.csv"   # the real export

cmap <- read_csv("choice-map.csv", show_col_types = FALSE)
```

> **Everything in this document runs on fake data until you change that line.**
> The numbers you see now were generated to match the published result. They are
> not our data and they are not evidence of anything.

## What we ran

Three boxes on every screen: two open, one covered. Each time you were asked for
the box matching a description. **If neither open box matches, the box you want
must be the closed one** — that is the whole logic of the paradigm.

Half of you got the **scalar** version (*some of the cookies*), half the
**number** version (*two fish*). That was between subjects, decided by a
randomiser.

The trials that carry the argument are the **critical** ones, where no open box
matches on the strengthened reading:

| | display | if you read it *exactly* / *not all* | if you read it loosely |
|---|---|---|---|
| scalar | target has **none** · target has **all** | neither matches → **covered** | the *all* box |
| number | **1** fish · **3 or 5** fish | neither is two → **covered** | the *more* box |

Huang, Spelke & Snedeker found **13%** covered in the scalar version and
**100%** in the number version. That asymmetry is the finding: *two* behaves as
though its upper bound is part of its meaning, and *some* does not — in a
population that computes scalar implicatures perfectly well elsewhere.

Everything else in the survey exists to make that comparison interpretable, and
each of those checks gets a section below.

## Reading the export

Qualtrics writes two metadata rows under the header before any data. Forgetting
them is the single most common way to get a nonsense result from a Qualtrics
export, because they parse as ordinary rows.

```{r read}
raw <- read_csv(data_file, col_types = cols(.default = col_character())) |>
  slice(-(1:2))

complete <- raw |> filter(Finished %in% c("True", "TRUE", "true", "1"))

cat("rows in file:", nrow(raw),
    "| completed:", nrow(complete),
    "| abandoned:", nrow(raw) - nrow(complete), "\n")
```

## Exclusions

Four familiarization trials: find a coloured shape. On two the shape is visible;
on two it is not, so the covered box is correct. Anyone who misses one was not
reading the instructions.

The rule is **all four correct**. That is strict, and it is a choice you should
be prepared to defend in the write-up — 3 of 4 is also defensible. Try both and
see whether it changes anything.

```{r exclusions}
correct <- c(fam1 = "1", fam2 = "1", fam3 = "3", fam4 = "3")

fam <- complete |>
  select(ResponseId, all_of(names(correct))) |>
  pivot_longer(-ResponseId, names_to = "trial", values_to = "resp") |>
  mutate(ok = resp == correct[trial]) |>
  group_by(ResponseId) |>
  summarise(n_correct = sum(ok, na.rm = TRUE), .groups = "drop") |>
  mutate(passed = n_correct == 4)

count(fam, n_correct, passed)
```

```{r excluded}
cat("excluded:", sum(!fam$passed), "of", nrow(fam), "\n")
keep <- fam$ResponseId[fam$passed]
```

## Getting to one row per trial

```{r tidy}
dat <- complete |>
  filter(ResponseId %in% keep) |>
  select(ResponseId, matches("^(scalar|number)_")) |>
  pivot_longer(-ResponseId, names_to = "q", values_to = "resp") |>
  filter(!is.na(resp), resp != "") |>
  separate_wider_delim(q, "_", names = c("term", "trial_type", "set"),
                       too_few = "align_start") |>
  mutate(set = replace_na(set, "s0"),
         participant = ResponseId,
         term = factor(term, levels = c("scalar", "number")),
         resp = as.integer(resp),
         covered = resp == 3,
         question = if_else(set == "s0", paste0(term, "_", trial_type),
                            paste0(term, "_", trial_type, "_", set))) |>
  left_join(cmap |> filter(choice_id != "correct") |>
              mutate(choice_id = as.integer(choice_id)),
            by = c("question", "resp" = "choice_id")) |>
  mutate(match_box = meaning == "match",
         critical  = trial_type %in% c("critical", "criticalOneSet"),
         probe     = trial_type %in% c("probe", "probeEarly")) |>
  select(participant, term, trial_type, set, resp, meaning,
         critical, probe, covered, match_box)

glimpse(dat)
```

A sanity check before any means. Everyone should have the same number of trials,
all in one term.

```{r sanity}
dat |> count(participant, term, name = "n_trials") |> count(term, n_trials)
```

## Did the task work at all?

Three separate checks, none of them about implicature. If these fail, nothing
downstream is interpretable.

### Huang et al.'s control trials

A match is visible on these, so essentially everyone should take it.

```{r controls}
HS <- c("noneSome", "someAll", "oneTwo", "twoMore")
dat |> filter(trial_type %in% HS) |>
  group_by(term, trial_type) |>
  summarise(chose_match = mean(match_box), n = n(), .groups = "drop")
```

### The shape fillers

These sit between the critical trials and contain no quantifier at all, so they
reset attention without priming *some*, *all* or *none*. One of the three
(`shape2`) needs the covered box.

```{r shapes}
dat |> filter(grepl("^shape", trial_type)) |>
  group_by(term, trial_type) |>
  summarise(correct = mean(if_else(trial_type == "shape2", covered, match_box)),
            n = n(), .groups = "drop")
```

### Was the covered box still a live option?

This one deserves care. **In none of the critical trials is the covered box
unambiguously correct** — it is always the diagnostic option. So a participant
who reads *some* loosely never needs it, and if it quietly stops feeling
available, a low covered-box rate is **extinction rather than semantics**.

We ask twice: once immediately before the critical trials, once after
everything. The second is the harder test, since by then every intervening
trial has had a visible answer.

```{r probes}
dat |> filter(probe) |>
  mutate(when = if_else(trial_type == "probeEarly",
                        "1 · before the critical trials",
                        "2 · after everything")) |>
  group_by(term, when) |>
  summarise(chose_covered = mean(covered), n = n(), .groups = "drop")
```

> **Discussion.** Suppose the *after* probe came back at 60%. What would you be
> able to say about the critical trials, and what would you not?

## Was the domain read box-internally?

*Give me the box where Zip has some of the apples* leaves open what **the
apples** ranges over: the apples in the box being considered, or all the apples
on screen?

It matters more than it sounds. On a standard critical trial the *all* box shows
the target with four of the eight apples visible — read globally, that **is**
"some but not all". So someone could take that box with the exclusive reading of
*some* completely intact, and the headline would say nothing about implicature.

The first trial asks who has **all** of the objects, which separates the
readings: box-internally the *all* box is correct, globally no box is.

```{r anchor}
anchor <- dat |> filter(trial_type %in% c("anchorAll", "anchorFive")) |>
  select(participant, term, box_internal = match_box)

anchor |> group_by(term) |>
  summarise(read_box_internally = mean(box_internal), n = n(), .groups = "drop")
```

## The main result

```{r critical-by-kind}
by_kind <- dat |> filter(critical) |>
  group_by(participant, term, trial_type) |>
  summarise(covered = mean(covered), .groups = "drop") |>
  group_by(term, trial_type) |>
  summarise(covered = mean(covered), n = n(), .groups = "drop")

by_kind |> mutate(design = if_else(trial_type == "critical",
                                   "Huang et al. — global reading available",
                                   "ours — domains coincide"))
```

The two rows for the scalar term differ in **one thing only**: whether the
global reading is available. `criticalOneSet` puts every object in a single box,
so the global set and the box-internal set are the same set and the loophole
closes. **The gap between them measures how much of the standard result the
global reading was buying.** If they agree, it was buying nothing.

```{r critical-pooled}
crit <- dat |> filter(critical) |>
  group_by(participant, term) |>
  summarise(covered = mean(covered), .groups = "drop")

crit |> group_by(term) |>
  summarise(mean = mean(covered), sd = sd(covered), n = n(),
            se = sd / sqrt(n),
            ci_low  = mean - qt(.975, n - 1) * se,
            ci_high = mean + qt(.975, n - 1) * se, .groups = "drop")
```

Published: scalar **.13**, number **1.00**.

```{r plot-critical}
theme_set(theme_minimal(base_size = 12))
pal <- c(scalar = "#4C72B0", number = "#CC0000")

summ <- crit |> group_by(term) |>
  summarise(mean = mean(covered), n = n(), se = sd(covered) / sqrt(n),
            lo = pmax(0, mean - 1.96 * se), hi = pmin(1, mean + 1.96 * se),
            .groups = "drop")

ggplot(crit, aes(term, covered, colour = term)) +
  geom_hline(yintercept = c(.13, 1), linetype = "dotted", colour = "grey55") +
  geom_jitter(width = .12, height = .015, alpha = .45, size = 2) +
  geom_pointrange(data = summ, aes(term, mean, ymin = lo, ymax = hi),
                  colour = "black", size = .7, inherit.aes = FALSE) +
  scale_colour_manual(values = pal, guide = "none") +
  scale_y_continuous(labels = scales::percent, limits = c(-.03, 1.03)) +
  labs(title = "Critical trials: no open box matches on the strengthened reading",
       subtitle = "Dotted lines are Huang, Spelke & Snedeker's 13% and 100%",
       x = NULL, y = "Chose the covered box")
```

```{r plot-choices}
dat |>
  mutate(choice = factor(resp, 1:3,
           labels = c("first open box", "second open box", "covered")),
         panel = case_when(critical ~ "critical trials",
                           probe    ~ "probe trials",
                           TRUE     ~ "everything else")) |>
  count(term, panel, choice) |>
  group_by(term, panel) |> mutate(p = n / sum(n)) |> ungroup() |>
  ggplot(aes(choice, p, fill = term)) +
  geom_col(position = "dodge") +
  facet_wrap(~panel) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Where the choices went", x = NULL, y = NULL, fill = NULL) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Is the difference real?

Term is **between** subjects, so this is a two-sample comparison on
participant-level rates. Huang et al. used Mann–Whitney, so we do too.

```{r test}
wilcox.test(covered ~ term, data = crit)
```

### Why there is no logistic model here

Try one and R will report an enormous coefficient with a *p*-value near 1, which
looks exactly like a null result and is its precise opposite.

If one condition sits at 0% or 100%, the groups are **completely separated**: no
finite log-odds describes the difference, so the estimate runs off to infinity
and the standard error goes with it. The effect is not too small to estimate. It
is too large.

```{r separation}
rates <- crit |> group_by(term) |> summarise(m = mean(covered), .groups = "drop")
if (any(rates$m %in% c(0, 1))) {
  cat("Complete separation:",
      paste0(rates$term, " at ", round(100 * rates$m), "%", collapse = ", "),
      "\nSkipping the logistic model — see above.\n")
} else {
  library(lme4)
  m <- glmer(covered ~ term + (1 | participant) + (1 | set),
             data = filter(dat, critical), family = binomial)
  print(summary(m)$coefficients)
}
```

> **Discussion.** Is "the effect is too large to model" a satisfying thing to
> write in a results section? What would you report instead?

## By items

Every critical trial used a different object. If the effect lives in one
picture rather than in the words, this is where it shows.

```{r by-items}
dat |> filter(critical) |>
  group_by(term, set) |>
  summarise(covered = mean(covered), n = n(), .groups = "drop") |>
  arrange(term, set)
```

## Language background

Optional, asked last. Huang et al. recruited English-speaking undergraduates, so
this is what makes the comparison with their rate a **check** rather than an
assumption.

Read the cell sizes before the rates. This is classroom data from people who
know each other, so report it in aggregate and do not push it further than it
goes.

```{r language}
lang <- complete |> filter(ResponseId %in% keep) |>
  transmute(participant = ResponseId,
            first_language = case_when(
              first_language == "1" ~ "English",
              first_language == "2" ~ "another language",
              first_language == "3" ~ "declined",
              TRUE ~ "no answer"))

count(lang, first_language)
```

```{r language-split}
crit |> left_join(lang, by = "participant") |>
  group_by(term, first_language) |>
  summarise(covered = mean(covered), n = n(), .groups = "drop")
```

## Writing it up

Which numbers belong where in the replication report:

| section | what goes in it |
|---|---|
| **Methods** | how many completed, how many excluded and on what rule, the split between terms, language background |
| **Results** | the control and shape-filler rates first (the task worked), then the probes and the anchor (the measure means what we think), then the critical rates and the test |
| **Discussion** | the comparison with 13% / 100%; what the `critical` vs `criticalOneSet` gap says about the domain; what the probes say about extinction; where our design departs from theirs and why |

Departures worth naming, because a reader will ask:

- Huang et al. put **trial type** between subjects as well as term. Ours is
  within, so everyone sees every trial type. At class sample sizes their design
  would give about five people per cell.
- We added the `criticalOneSet` condition, the anchor, two probes and the shape
  fillers. None of these is in the original.
- Their familiarization ran **twice**, with feedback on the first pass. Ours runs
  once with none, because Qualtrics cannot easily give feedback — which is part
  of why the probes are there.

> **A caveat worth quoting.** Lyn Frazier, on this literature: *"studies like
> those reported here do not necessarily provide an accurate estimate of the
> absolute number of implicatures drawn. Rather, it is the relation between
> conditions that may be interpreted securely."* Our absolute rates are soft.
> The difference between scalar and number is not.

## Looking ahead

We have just spent a session asking whether a difference between two conditions
is bigger than chance. In two weeks we start asking a different question of the
same kind of data: not *is there an effect* but *does a model predict the whole
pattern, and does it beat a model without the piece we care about*.

Those are genuinely different standards of evidence, and neither subsumes the
other. Worth holding onto the feeling of this session when RSA arrives.

## Your turn

1. Re-run the exclusion rule at 3-of-4 instead of 4-of-4. Does anything change?
2. Compute the critical rate using only each participant's **first** critical
   trial. Should it differ from the average over all four? Does it?
3. Split the critical rate by the anchor response. What would you conclude if
   the people who read the domain globally behaved differently?
4. The number condition is at 100%, so its confidence interval has zero width.
   Is that honest? What would you put on the plot instead?
